### 2023-09-06 doc/15yrsChange.Rmd
incorporate references

# Papers to read
- earth system models (pay special attention to predictions and uncertainty for eastern US)
  - @wu_uncertainty_2023. - do they say anything about nuisance species?
  - @ahlstrom_robustness_2012
- @miller_overabundant_2023
  - what were there metrics? would it be worth applying their metrics to our forest in order to classify by their system?

@@

The temperate forest biome plays a critical role in the global carbon cycle, accounting for almost half of the net global forest carbon (C) sink [@harris_global_2021], with deciduous forests representing a substantial portion of this, sequestering >300 Tg C yr^-1^ [@pugh_role_2019]. 
Although currently a C sink, the future of the biome remains uncertain.
The dominant view imbued in Earth System Models is that, at least within the eastern US, the biome is likely to remain a C sink for the remainder of the 21st century [@finzi_carbon_2020; @wu_uncertainty_2023], albeit with declining CO~2~ sequestration capacity [@ahlström_robustness_2012]. 
Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlstrom_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014].
One mechanism that is not represented in global C models is the impact of nuisance species -- i.e., endemic or non-endemic species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in undesirable ecological consequences [@refs]. 
Similarly, the impact of nuisance species on forest carbon budgets is not considered in machine learning/niche models that seek to project future forest distribution and carbon stocks [**?** @wu_uncertainty_2023], nor in carbon offset projects [**???** @refs].
This is problematic in that nuisance species -- including non-endemic insect pests and pathogens, non-endemic plants, and over-abundant herbivores -- are dramatically impacting carbon cycling in temperate deciduous forests around the world [@refs].

@@

@@

@mcgarvey_effects_2013

>>>>

@@

The temperate forest biome plays a critical role in the global carbon cycle, accounting for almost half of the net global forest carbon (C) sink [@harris_global_2021], with deciduous forests representing a substantial portion of this, sequestering >300 Tg C yr^-1^ [@pugh_role_2019]. 
Although currently a C sink, the future of the biome remains uncertain.
The dominant view imbued in Earth System Models is that, at least within the eastern US, the biome is likely to remain a C sink for the remainder of the 21st century [@finzi_carbon_2020; @wu_uncertainty_2023], albeit with declining CO~2~ sequestration capacity [@ahlström_robustness_2012]. 
Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlström_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014].
One mechanism that is not represented in global C models is the impact of nuisance species -- i.e., endemic or non-endemic species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in undesirable ecological consequences [@refs]. 
Similarly, the impact of nuisance species on forest carbon budgets is not considered in machine learning/niche models that seek to project future forest distribution and carbon stocks [**?** @wu_uncertainty_2023], nor in carbon offset projects [**???** @refs].
This is problematic in that nuisance species -- including non-endemic insect pests and pathogens, non-endemic plants, and over-abundant herbivores -- are dramatically impacting carbon cycling in temperate deciduous forests around the world [@refs].

@@

@fei_biomass_2019

@@

@mcgarvey_effects_2013
@holm_interactive_2013
